Birth Spacing in the Presence of Son Preference and Sex-Selective \\ Abortions: India's Experience over Four Decades

### 2018-04-19 presentations/2018-paa-poster.tex
Add text

\documentclass[final]{beamer}
  \mode<presentation>
  {
  \usetheme{CCP}
  }
  \usepackage[orientation=landscape,scale=1.4]{beamerposter}

@@

\title{Birth Spacing in the Presence of Son Preference and Sex-Selective \\ Abortions: India's Experience over Four Decades}
\author{Claus P\"ortner}
\institute{Albers School of Business and Economics, Seattle University, \& Center for Studies in Demography and Ecology, University of Washington}
\footer{\texttt{cportner@seattleu.edu}}
\begin{document}

@@

\begin{column}{.3\linewidth}

\begin{block}{\LARGE Background}

The central question addressed here is the extent to which spacing patterns 
significantly changed as prenatal sex determination became available.
Specifically, did the spacing after girls increase relative to the spacing
after boys?
The underlying idea is that in the absence of sex selection son 
preference leads to shorter spacing after the birth of a girl than after the 
birth of a boy, whereas son preference increases the spacing after the birth 
of a girl relative to after the birth of a boy when prenatal sex selection is 
available.
I introduce a new method that simultaneously accounts for spacing
between births and the potential use of sex selection. 
I apply the method to over four decades of data from India's NFHS.

@@

\begin{block}{Objectives}

@@

\begin{column}{.3\linewidth}

@@

\begin{column}{.3\linewidth}

@@

The results show two very different approaches to son preference.

At one extreme, women without education mostly follow the standard 
pattern of shorter spacing when a woman does not have the desired number of 
sons.

At the other extreme, women with eight or more years of education have
experienced a reversal of the traditional spacing patterns.
Women with either no or one son now have substantially longer spacing 
than if they have two or more sons.

Although it is not possible to directly observe whether sex selection
causes the unequal sex ratio and longer spacing the results are
consistent with a general increase in sex selection over time.

Compared to previous research one result stands out: 
the use of sex selection is spreading to women with lower levels of education.

@@

>>>>

\documentclass[final]{beamer}
  \mode<presentation>
  {
  \usetheme{CCP}
  }
  \usepackage[orientation=landscape,scale=1.2]{beamerposter}

@@

\newlength{\sepwid}
\newlength{\onecolwid}
\newlength{\twocolwid}
\newlength{\threecolwid}
\setlength{\sepwid}{0.024\paperwidth}
\setlength{\onecolwid}{0.24\paperwidth}
\setlength{\twocolwid}{0.4\paperwidth}
\setlength{\threecolwid}{0.19\paperwidth}

\title{Birth Spacing in the Presence of Son Preference and Sex-Selective \\ Abortions: India's Experience over Four Decades}
\author{Claus P\"ortner}
\institute{Albers School of Business and Economics, Seattle University, \& Center for Studies in Demography and Ecology, University of Washington}

%%% Put the name of conference here.
	\def\conference{2018 Annual Meeting of the Population Association of America} 
 %%% Put your e-mail address here.
 	\def\yourEmail{cportner@seattleu.edu}

\begin{document}

@@

\begin{column}{\onecolwid}

\begin{block}{Background}

Son preference is often associated with shorter spacing after the 
births of girls than boys, shorter spacing is, in turn, associated 
with worse health outcomes for girls and mothers.

However, the introduction of prenatal sex determination fundamentally changed the 
relationship between son preference and birth spacing.
Couples, who had shorter birth spacing before because of son preference,
now have access to prenatal sex determination and sex-selective abortions,
and each abortion increases birth spacing by six months to a year.

As a result, we now have a situation where we may observer \emph{longer} spacing
for families with daughters than for families with sons, precisely because
strong son preference leads them to use sex selection.

To further complicate matters, we can still observe shorter birth spacing 
after the births of daughters as a representation of son preference for 
families who---for one reason or another---do not use prenatal sex selection.

@@

\begin{alertblock}{Contributions}

\begin{description}

\item[Main idea:]  in the absence of sex selection son preference leads to shorter 
spacing after the birth of a girl than after the birth of a boy, whereas 
son preference increases the spacing after the birth of a girl relative 
to after the birth of a boy when prenatal sex selection is available.

\item[Method:] I introduce an empirical method that directly incorporates 
the effects of sex-selective abortions on 
the duration between births
\emph{and} 
the likelihood of a son.

\item[Central question:] how spacing patterns changed with access
to prenatal sex determination became available.
Specifically, did the spacing after girls increase relative to the spacing
after boys?
\end{description}

\end{alertblock}

@@

\begin{column}{\twocolwid}

@@

\begin{column}{\onecolwid}

@@

\begin{enumerate}
\item there has been a general increase in the length of spacing between births
over the four decades covered by the data.
The exception is for first births, where the median duration has either
remained the same or slightly declined, although this hides a
significant compression of the variation in spell length.

\item the most substantial increase in spacing is for the women who
are most likely to use sex selection.
Among the best-educated women, those with no boys now has 
significantly longer spacing---and a more male-biased sex ratio---than 
similar women with boys.

\item women with no education still follow the standard pattern of
short spacing when they have girls and little evidence of the use of sex
selection.
In other words, these women adhere to a strategy where they achieve a
son through higher fertility rather than the use of sex selection.

\item sex ratios are the most likely to decline within spells at
lower-order spells, where the pressure to provide a son is smaller, and
are more likely to increase or remain consistently high for higher-order
spells, where the pressure to ensure a son is high.

\end{enumerate}

@@

\begin{block}{Acknowledgements}

Support for development of the method from the University of Washington Royalty 
Research Fund and the Development Research Group of the World Bank is gratefully 
acknowledged.
The views and findings expressed here are those of the author and
should not be attributed to the World Bank or any of its member countries.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
